From Reversible Programs to Univalent Universes and Back

### 2017-05-02 upi.tex
removed obsolete fwhott
added many 2 combinators

%% Amr
%% words to remember :-)
%% sublime unfathomable
%% ---

% \usepackage{fdsymbol}
\usepackage{bbold}
\usepackage{bussproofs}
\usepackage{keystroke}
\usepackage{comment}
\usepackage{tikz}

%% \usepackage{hyphenat}
%% \usepackage{color}
%% \usepackage{url}
%% \usepackage{amssymb}
%% \usepackage{amsmath}
%% \usepackage{mdwtab}
%% \usepackage{mdwlist}
%% \usepackage{listings}
%% \lstMakeShortInline[columns=fullflexible]|
%% \lstnewenvironment{code}{\lstset{basicstyle={\sffamily\footnotesize}}}{}

@@

\title[Featherweight HoTT]{Univalent Universes and Completness of Reversible Programs: \\
  Featherweight HoTT}

\author{X}
\affiliation{
  \institution{Y}
  \country{Z}}
\email{A@B.C}

\begin{abstract}
\end{abstract}

@@

Our conjecture:

@@

Following the HoTT book and the slides on ``A Characterization of Univalent
Fibrations'' by Dan Christensen.

@@

\paragraph*{Equivalences (Chapter 4 in the book)} For any two types $A$ and $B$
we can form the type $A \simeq B = \Sigma_{(f : A \to B)} \mathit{isequiv}~f$ of
equivalences. Each equivalence consists of a function $f : A \to B$ together
with evidence that this function is an equivalence.

@@

\item For any type $F$ we define
  $\{F\} = \Sigma_{A : \mathcal{U}} \| F \simeq A \|$. By a proposition of
  Christensen, any fibration with base space $\{F\}$, i.e., any fibration
  $\{F\} \to \mathcal{U}$, is univalent. In particular, we can take the base
  space to be:
  \begin{itemize}
  \item $\{\ot\}$. The type $\{\ot\}$ is actually the same as $\ot$ and this was
    covered by previous examples.
  \item $\{S^1\}$. The type $\{S^1\}$ is the same as $S^1$. This is perhaps the
    simplest base space to choose but it does not naturally correspond to a
    conventional programming language.
  \item $\{\bt\}$. In some sense this is the simplest
    programming-language--relevant universe --- our featherweight HoTT. We aim
    to characterize this base space as a reversible programming language based
    on type isomorphisms.
  \end{itemize}

@@

%%%%%
\subsection{$\{\bt\}$}

Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$.

%%%%%
\subsection{$\Pi$}

@@

We begin by unpacking the objects, 1-paths, and 2-paths of $\{\bt\}$.

>>>>

%% Amr
%% words to remember :-)
%% sublime unfathomable
%% path categorical semantics
%% ---

\usepackage{bbold}
\usepackage{bussproofs}
\usepackage{keystroke}
\usepackage{comment}
\usepackage{tikz}

@@

\title{From Reversible Programs to Univalent Universes and Back}
\author{Jacques Carette}
\affiliation{
  \institution{McMaster University}
  \country{Canada}}
  \email{carette@mcmaster.ca}
\author{Chao-Hong Chen}
\author{Vikraman Choudhury}
\author{Robert Rose}
\author{Amr Sabry}
\affiliation{
  \institution{Indiana University}
  \country{United States}}
\email{\{chen464,vikraman,rrose1,sabry\}@indiana.edu}

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a univalent (in the sense of Homotopy Type
  Theory) universe. The correspondence relates combinators witnessing type
  isomorphisms in the programming language to paths in the univalent universe,
  and combinator optimizations in the programming language to 2-paths in the
  univalent universe. The result provides a simple computational interpretation
  of paths, 2-paths, and univalence in terms of simple programming
  constructs.
\end{abstract}

@@

We propose a featherweight very specialized version of HoTT: a smaller
collection of types and only reversible functions. No dependent types!  More
precisely, our conjecture:

@@

Following the HoTT book and the slides on ``A Characterization of Univalent
Fibrations'' by Dan Christensen. All the definitions and most of the examples in
this section are formalized in the accompanying Agda code.

@@

\paragraph*{Equivalences (Chapter 4 in the book)} For any two types $A$ and $B$
we can form the type $A \simeq B = \Sigma_{(f : A \to B)} \mathit{isequiv}~f$ of
equivalences. Each equivalence consists of a function $f : A \to B$ together
with evidence $\mathit{equiv}~f$ that this function is an equivalence.

@@

\item For any type $F$ we define
  $\{F\} = \Sigma_{(A : \mathcal{U})} \| F \simeq A \|$. By a proposition of
  Christensen, any fibration with base space $\{F\}$, i.e., any fibration
  $\{F\} \to \mathcal{U}$, is univalent. In particular, we can take the base
  space to be:
  \begin{itemize}
  \item $\{\ot\}$. The type $\{\ot\}$ is actually the same as $\ot$ and this was
    covered by previous examples.
  \item $\{S^1\}$. The type $\{S^1\}$ is the same as $S^1$. This is perhaps the
    simplest base space to choose but it does not naturally correspond to a
    conventional programming language.
  \item $\{\bt\}$. In some sense this is the simplest
    programming-language--relevant universe --- our featherweight HoTT. We aim
    to characterize this base space as a reversible programming language based
    on type isomorphisms.
  \end{itemize}

@@

%%%%%
\subsection{The Subuniverse $\{\bt\}$}

Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$. We begin by
unpacking the objects, 1-paths, and 2-paths of $\{\bt\}$.

%%%%%
\subsection{The Language $\Pi$ of Type Isomorphisms}

@@
